---
title: "Reproducible Research Course Project 1"
output: html_document, md_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading Data and Preprocessing

Data is read from a local folder, and the date is converted to a POSIXct format.

``` {r Preprocessing}
library(lubridate)
activity <- read.csv("./activity.csv", header = TRUE)
activity$date <- ymd(activity$date)
```

## Totals, Mean and Median each day

Totals are calculated for each day and a histogram is plotted using the base plotting system.

``` {r}
library(dplyr)
bydays <- group_by(activity[!is.na(activity$steps),], date)
totals <- summarize(bydays, total_steps = sum(steps, na.rm = TRUE))
hist(totals$total_steps, xlab = "Steps Per Day", main = "Frequency of Total Steps Per Day")

mean_steps <- mean(totals$total_steps)
median_steps <- median(totals$total_steps)

```

The mean number of steps each day is `r mean_steps` and the median number of steps each day is `r median_steps`.

## Average Daily Activity Pattern

```{r Intervals}
byinterval <- group_by(activity, interval)
intervalmeans <- summarize(byinterval, mean_steps = mean(steps, na.rm = TRUE))
with(intervalmeans, plot(interval, mean_steps, xlab = "Time Interval",
                         ylab = "Mean Steps", 
                         main = "Mean Steps Per Time Interval for All Days"))
maxtime <- intervalmeans$interval[which.max(intervalmeans$mean_steps)]
```

The interval with the largest mean number of steps is `r maxtime`.

## Imputing missing values

To impute missing values, each missing value will be replaced with the mean for that time interval, as calculated in the above intervalmeans table.

```{r Impute Values}
library(plyr)
missing <- sum(as.numeric(is.na(activity$steps)))
new_activity <- ddply(activity, .(interval), mutate, intervalmeans = mean(steps, na.rm = TRUE))
new_activity$steps <- ifelse(is.na(new_activity$steps), new_activity$intervalmeans, new_activity$steps) 
new_totals = new_activity %>% group_by(date) %>% summarise_at(c("steps"), funs(sum))
hist(new_totals$steps, xlab = "Steps Per Day", main = "Frequency of Total Steps Per Day")
new_mean_steps <- mean(totals$total_steps)
new_median_steps <- median(totals$total_steps)

```

There are `r missing` missing values in the activity data set. 

The mean number of steps each day is `r new_mean_steps` and the median number of steps each day is `r new_median_steps`. These are not substantially different from the mean and median with NA values removed.

## Weekdays Versus Weekends

The graphs below show the difference in the mean steps per 5-minute interval averaged over all weekdays (top) and weekends (bottom).

``` {r Weekday versus Weekend}

new_activity <- mutate(new_activity, daytype = factor(1*(weekdays(date) %in% c("Sunday", "Saturday")), labels = c("weekday", "weekend")))

weekdays <- subset(new_activity, daytype == "weekday")
weekends <- subset(new_activity, daytype == "weekend")

weekday_byinterval <- group_by(weekdays, interval)
weekend_byinterval <- group_by(weekends, interval)

weekday_intervalmeans <- weekday_byinterval %>% group_by(interval) %>%
        summarise_at(c("steps"), funs(mean))
weekend_intervalmeans <- weekend_byinterval %>% group_by(interval) %>%
        summarise_at(c("steps"), funs(mean))

par(mfrow = c(2, 1), mar = c(5,4,2,1))
with(weekday_intervalmeans, plot(interval, steps, type = "l", main = "Weekday",
                                 ylab = "Mean Steps", xlab = "Interval"))
with(weekend_intervalmeans, plot(interval, steps, type = "l", main = "Weekend",
                                 ylab = "Mean Steps", xlab = "Interval"))
```
